---
title: "Calculating contribution to annual rate"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

mm23 contains functions to reproduce the calculation of an items contribution to
its parents annual inflation rate. These are quite niche, and rely on the data
being in a specific structure. *Theres no real error checking in here (yet) so
take care!*

This article will explain the steps needed to reproduce the calculation. 

## Weights

CPIH weights are a problem. They are adjusted twice a year (since 2017), in
December and January. But the mm23 dataset only has a single weight for each
year. The only place the full weights are exposed are in the 'detailed reference
tables' spreadsheet. The functions `get_cpi*_weights()` which generates a
weights dataset. *It may not be very robust, it needs some testing, but if
everything is in the right place it works OK.*

It returns a dataset with a weight for every month.

## Unchaining and calculating contribution

The `unchain()` function can be used within a `mutate()` call to produce an
unchained index. Its based on the assumption that CPI weights change twice a
year, and so an unchained index is based on previous January = 100, or for the
January index, based on previous December = 100. Example code below.

## Calculating contribution to 'all items' 12 month rate

The calculation is as follows, where:

$c = component\ c$

$a =\ 'all\ items'\ CPI\ index$

$W^c_{y} = weight\ of\ component\ c\ in\ year\ y$
$I^c_t = index\ for\ component\ c\ in\ month\ t\ based\ on\ January\ of\ current\ year =100$
$I^a_{Jan} = all\ items\ index\ for\ January\ based\ on\ previous\ month\ (December) = 100$
$I^a_{Dec} = all\ items\ index\ for\ December\ based\ on\ previous\ January = 100$



$$
(\frac{W^c_{y-1}} {W^a_{y-1}})
\times
(\frac{({I^c_{Dec}} - {I^c_{t-12}})}{I^a_{t-12}})
\times
100
\ \ +\ \ 
(\frac{W^c_y}{W^a_y})
\times
(\frac{({I^c_{Jan}}-100)}{I^a_{t-12}})
\times
I^a_{Dec}
\ \ +\ \
(\frac{W^c_y}{W^a_y})
\times
(\frac{(I^c_t - 100)}{I^a_{t-12}})
\times
\frac{I^a_{Jan}}{100}
\times
I^a_{Dec}
$$


Note that 'all items' doesn't have to be all items CPIH, it can be a category
total (such as food). It is important that the calculations are performed using
**unchained indices** (i.e. based on previous January = 100, or for the January
index, based on previous December =100). For the month of interest, the
contribution of each component of the CPI to the 12-month rate is calculated.
The same is done for the preceding month. The differences between the two are
the contributions to the *change* in the CPI 12-month rate.


The `contribution()` function applies this calculation to a wide dataset (1
column with month number or date, and the indices and weights in subsequent
columns). It can then be used in a `mutate()` call. It takes $I^a$, $W^a$,
$I^c$, and $W^c$ as parameters, and leverages `dplyr::lag()` to do the rest.
Example below.

## Contribution - data setup

Lets get the data we need for the calculations and reduce the need for lots of
downloads.

```{r data_setup, message=FALSE}
library(mm23)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)

# Get our data together
mm23 <- acquire_mm23()
data <- get_mm23_month(mm23)
cpih_wts <- get_cpih_weights()
cpi_wts <- get_cpi_weights()

```


## CPIH food weight contribution

Code to calculate the contribution to changes in food and non-alcoholic drinks
inflation.


```{r cpih_contribution, message=FALSE}

# Select and unchain food indices
food_cdids <- c("L523", "L52I", "L52J", "L52K",
                "L52L", "L52M", "L52N", "L52O",
                "L52P", "L52Q", "L52S", "L52T")


food_unchained <- data |> 
  filter(cdid %in% food_cdids & date >= "2017-01-01") |> 
  group_by(cdid) |> 
  mutate(unchained_value = unchain(month(date), value)) |> 
  select(date, cdid, value = unchained_value)


# Extract the relevant food weights
food_weights <- c("L5CZ", "L5DH", "L5DI", "L5DJ",
                  "L5DK", "L5DL", "L5DM", "L5DN",
                  "L5DO", "L5DP", "L5DR", "L5DS")


foodwts <- cpih_wts |> 
  filter(cdid %in% food_weights & date >= "2017-01-01" & date <= "2023-03-01") 


# Combine the indices and weights and calculate the contribution to annual rate
unchained <- food_unchained |> 
  bind_rows(foodwts) |> 
  pivot_wider(names_from = cdid)

contribution <- unchained |> 
            mutate(bread = contribution(month = month(date),
                                        all_items_index = L523,
                                        all_items_weight = L5CZ,
                                        component_index = L52I,
                                        component_weight = L5DH),
                   meat = contribution(month = month(date),
                                        all_items_index = L523,
                                        all_items_weight = L5CZ,
                                        component_index = L52J,
                                        component_weight = L5DI),
                   fish = contribution(month = month(date),
                                        all_items_index = L523,
                                        all_items_weight = L5CZ,
                                        component_index = L52K,
                                        component_weight = L5DJ),
                   dairy = contribution(month = month(date),
                                        all_items_index = L523,
                                        all_items_weight = L5CZ,
                                        component_index = L52L,
                                        component_weight = L5DK),
                   oils = contribution(month = month(date),
                                        all_items_index = L523,
                                        all_items_weight = L5CZ,
                                        component_index = L52M,
                                        component_weight = L5DL),
                   fruit = contribution(month = month(date),
                                        all_items_index = L523,
                                        all_items_weight = L5CZ,
                                        component_index = L52N,
                                        component_weight = L5DM),
                   vegetables = contribution(month = month(date),
                                        all_items_index = L523,
                                        all_items_weight = L5CZ,
                                        component_index = L52O,
                                        component_weight = L5DN),
                   sugar = contribution(month = month(date),
                                        all_items_index = L523,
                                        all_items_weight = L5CZ,
                                        component_index = L52P,
                                        component_weight = L5DO),
                   food_nec = contribution(month = month(date),
                                        all_items_index = L523,
                                        all_items_weight = L5CZ,
                                        component_index = L52Q,
                                        component_weight = L5DP),
                   coffee = contribution(month = month(date),
                                        all_items_index = L523,
                                        all_items_weight = L5CZ,
                                        component_index = L52S,
                                        component_weight = L5DR),
                   water = contribution(month = month(date),
                                        all_items_index = L523,
                                        all_items_weight = L5CZ,
                                        component_index = L52T,
                                        component_weight = L5DS)
                   )

head(contribution) |> knitr::kable()

# put the data back into tidy form and plot some results
cont <- contribution |> 
  select(date, bread:water) |> 
  pivot_longer(cols = bread:water) |> 
  filter(!is.na(value))

cont |> 
  filter(date >= "2021-07-01") |> 
  ggplot() +
  geom_col(aes(x = date, y = value, fill = name)) +
  labs(title = "Contribution to annual food and non-alc inflation rate")

```


## CPI food weight ontribution
```{r cpi_contribution, message=FALSE}
# Select and unchain food indices
food_cdids <- c("D7BU",
                "D7D5",
                "D7D6",
                "D7D7",
                "D7D8",
                "D7D9",
                "D7DA",
                "D7DB",
                "D7DC",
                "D7DD",
                "D7DE",
                "D7DF")


food_unchained <- data |> 
  filter(cdid %in% food_cdids & date >= "2017-01-01") |> 
  group_by(cdid) |> 
  mutate(
    unchained_value = unchain(month(date), value)
  ) |> 
  select(date, cdid, value = unchained_value)


# Extract the relevant food weights
food_weights <- c("CHZR",
                  "CJWB",
                  "CJWC",
                  "CJWD",
                  "CJWE",
                  "CJWF",
                  "CJWG",
                  "CJWH",
                  "CJWI",
                  "CJWJ",
                  "CJWK",
                  "CJWL")


foodwts <- cpi_wts |> 
  filter(cdid %in% food_weights & date >= "2017-01-01" & date <= "2023-03-01") 


# Combine the indices and weights and calculate the contribution to annual rate
unchained <- food_unchained |> 
  bind_rows(foodwts) |> 
  pivot_wider(names_from = cdid)

contribution <- unchained |> 
  mutate(`Bread and cereals` = contribution(month = month(date),
                              all_items_index = D7BU,
                              all_items_weight = CHZR,
                              component_index = D7D5,
                              component_weight = CJWB),
         `Meat` = contribution(month = month(date),
                             all_items_index = D7BU,
                             all_items_weight = CHZR,
                             component_index = D7D6,
                             component_weight = CJWC),
         `Fish` = contribution(month = month(date),
                             all_items_index = D7BU,
                             all_items_weight = CHZR,
                             component_index = D7D7,
                             component_weight = CJWD),
         `Milk, cheese and eggs` = contribution(month = month(date),
                              all_items_index = D7BU,
                              all_items_weight = CHZR,
                              component_index = D7D8,
                              component_weight = CJWE),
         `Oils and fats` = contribution(month = month(date),
                             all_items_index = D7BU,
                             all_items_weight = CHZR,
                             component_index = D7D9,
                             component_weight = CJWF),
         `Fruit` = contribution(month = month(date),
                              all_items_index = D7BU,
                              all_items_weight = CHZR,
                              component_index = D7DA,
                              component_weight = CJWG),
         `Vegetables` = contribution(month = month(date),
                                   all_items_index = D7BU,
                                   all_items_weight = CHZR,
                                   component_index = D7DB,
                                   component_weight = CJWH),
         `Sugar, jam and confectionery` = contribution(month = month(date),
                              all_items_index = D7BU,
                              all_items_weight = CHZR,
                              component_index = D7DC,
                              component_weight = CJWI),
         `Food products (nec)` = contribution(month = month(date),
                                 all_items_index = D7BU,
                                 all_items_weight = CHZR,
                                 component_index = D7DD,
                                 component_weight = CJWJ),
         `Coffee and tea` = contribution(month = month(date),
                               all_items_index = D7BU,
                               all_items_weight = CHZR,
                               component_index = D7DE,
                               component_weight = CJWK),
         `Mineral waters, soft drinks` = contribution(month = month(date),
                              all_items_index = D7BU,
                              all_items_weight = CHZR,
                              component_index = D7DF,
                              component_weight = CJWL)
  )


# put the data back into tidy form and plot some results
cont <- contribution |> 
  select(date, `Bread and cereals`:`Mineral waters, soft drinks`) |> 
  pivot_longer(cols = `Bread and cereals`:`Mineral waters, soft drinks`) |> 
  filter(!is.na(value))

cont |> 
  filter(date >= "2020-01-01") |> 
  ggplot() +
  geom_col(aes(x = date, y = value, fill = name)) +
  labs(title = "Contribution to annual food and non-alc inflation rate")

```


