---
title: "Getting started with mm23"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Get MM23 data

Download the latest MM23 from ONS. `acquire_mm23()` simply downloads the full
latest `mm23.csv` file, and returns its location. By default it puts it into a
temporary file, but you can specify where you want to put it.

```{r}
library(mm23)

# Puts the data in a temporary file and returns its name
mm23 <- acquire_mm23()
mm23

# Or specify a location
# mm23 <- acquire_mm23("~/data")
# data <- read.csv("~/data/mm23.csv")

```

However, the raw `mm23.csv` file is not very user-friendly. The key metadata is
in the first few rows. Yearly, quarterly and monthly series values are included
in the csv file in the same column. Its hard to unpick the data. There are
thousands of series all included in a single csv file.

Which is mainly where `mm23` comes in.


## Wrangling MM23 with `mm23`

### Data

The `get_mm23_*` functions will return data in a tidy format. Each series has a
unique `CDID` identifier.

```{r message=FALSE, warning=TRUE, echo=FALSE}

mm23::get_mm23_month(mm23) |> head() |> knitr::kable()

```


Its more efficient to use `acquire_mm23` first so as to only download the data
once. But if used without parameters the `get_mm23_*` functions will download
the latest data.


```r
mm23 <- acquire_mm23()
m <- get_mm23_month(mm23)
q <- get_mm23_quarter(mm23)
y <- get_mm23_year(mm23)

```

### Metadata

Use `get_mm23_metadata()` to return details of what each series CDID represents.
Below is a random sample of its output. For some series `get_mm23_metadata()`
extracts the category of data included and puts it in the `category` field.
Equally, it infers the COICOP digit level and adds that to the `level` field.

```{r message=FALSE, warning=TRUE, echo=TRUE}

mm23::get_mm23_metadata(mm23) |> dplyr::slice_sample(n = 5) |> knitr::kable()

```




You can join data and metadata by CDID, for analysis.

```{r, message=FALSE}

data <- get_mm23_month(mm23)
metadata <- get_mm23_metadata(mm23)

data |>
  dplyr::filter(cdid %in% c("L55O", "L55P")) |>
  dplyr::left_join(metadata) |> 
  dplyr::select(date, cdid, value, title, category) |> 
  head() |> 
  knitr::kable()


data |>
  dplyr::filter(cdid %in% c("L55O", "L55P")) |>
  dplyr::left_join(metadata) |> 
  dplyr::select(date, cdid, value, title, category) |> 
  ggplot2::ggplot() +
  ggplot2::geom_line(ggplot2::aes(x = date, y = value, colour = title)) +
  ggplot2::theme(legend.position = "bottom")



```
