---
title: "CPI Price Quotes"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(mm23)

```

## Introduction

The price quote functions will generally access and download the price quote
data, but changes to the variables over the years and the nature of the data
means that you'll probably have to do some wrangling to get what you need -
assuming what you're interested in is the average retail price of stuff. The
relevant functions are:

 * `get_cpi_price_quotes()` for data from 2020 onwards, and
 * `get_cpi_price_quotes_archive()` for data from 2010-2019
 
The data is accessed from the url stored in `cpi_price_quotes_url()`. The
addresses of data 2010-2019 are stored in `archive_urls()`.

The data is returned as is - in order to look at prices we need to exclude
invalid records - pre February 2025 this is ones with a `validity` not 3 or 4,
post February 2025 `validity is FALSE.

## Sample code

#### Historic data

Lets look at some old bread prices.

```{r}
library(ggplot2)

y <- 2013:2014

bread <- purrr::map(y, \(y) {
  get_cpi_price_quotes_archive(y, foodonly = TRUE) |> 
    dplyr::filter(item_id %in% c(210102, 210111))
}) |> dplyr::bind_rows()

bread |> 
  dplyr::filter(validity %in% c(3,4)) |> 
  dplyr::group_by(lubridate::year(quote_date), item_desc) |>
  dplyr::summarise(avg = mean(price)) |> 
  knitr::kable()

```

### Recent data

We can compare with some more recent data:

```{r}

# If you want to get more than 1 year you can do eg rep(2024:2025, 12) and
# rep(1:12, 2). But you don't necessarily need purrr::map2
y <- rep(2024)
m <- rep(1:12)

more_bread <- purrr::map2(y, m, \(y,m) {
  get_cpi_price_quotes(y, m, foodonly = FALSE) |> 
    dplyr::filter(item_id %in% c(210102, 210111))
}) |> dplyr::bind_rows()

more_bread |> 
    dplyr::filter(validity %in% c(3,4)) |> 
  dplyr::group_by(quote_date, item_desc) |>
  dplyr::summarise(avg = mean(price)) |> 
  ggplot2::ggplot() +
  ggplot2::geom_line(ggplot2::aes(x = quote_date, y = avg, colour = item_desc))

```

The validity variable changed in Feb 2025:

```{r}

# Doing the same as more_bread, but without purrr::map2
latest_bread <- get_cpi_price_quotes(2025, 1:4, foodonly = FALSE) |> 
  dplyr::filter(item_id %in% c(210102, 210111)) |> 
  dplyr::mutate(valid = dplyr::case_when(quote_date <= "2025-01-01" & validity %in% c(3,4) ~ "Y",
                                  quote_date >= "2025-02-01" & validity == 1 ~ "Y",
                                  .default = "N"))

latest_bread |> 
  dplyr::filter(valid == "Y", quote_date >= "2013-01-01") |>
  dplyr::group_by(quote_date, item_desc) |> 
  dplyr::summarise(avg_price = mean(price)) |> 
  knitr::kable()

```

## Sample data

### Latest (post Jan 2025) data

```{r}
mm23::get_cpi_price_quotes(2025, 3) |> dplyr::slice_sample(n = 5) |> t() |> knitr::kable()
```

### Pre Feb 2025 data

```{r}
mm23::get_cpi_price_quotes(2021, 6) |> dplyr::slice_sample(n = 5) |> t() |> knitr::kable()
```



## Raw data formats

### Raw data format (Feb 2025 onwards)

| COLUMN HEADING         | DESCRIPTION                                                                                                                             |
|------------------------|-----------------------------------------------------------------------------------------------------------------------------------------|
| QUOTE_DATE             | The year and month in which the individual price observation was collected.                                                             |
| CS_ID                  | Unique identification number of the Consumption Segment or Item collected. Value is equal to ITEM_ID if no CS alternative is available. |
| CS_DESC                | Description of the Consumption Segment or Item collected.                                                                               |
| ITEM_ID                | Unique identification number of the Item collected.                                                                                     |
| ITEM_DESC              | Description of the Item collected.                                                                                                      |
| VALIDITY               | The status of the individual price quote:                                                                                               |
|                        | TRUE = Quote included in the current months index.                                                                                      |
|                        | FALSE = Quote not included in the current months index.                                                                                 |
| SHOP_CODE              | Shop code from which the individual price quote was obtained.                                                                           |
| PRICE                  | Observed price on date of collection (in £GBP).                                                                                         |
| INDICATOR_BOX          | Indicator used to highlight any item or price change:                                                                                   |
|                        | C = Comparable – change in product being priced, new product is similar to the previous product.                                        |
|                        | M = Missing – not sold at shop.                                                                                                         |
|                        | N = Non-comparable – change in product being priced, new product is not comparable to the previous product.                             |
|                        | P = Current price unavailable for the period.                                                                                           |
|                        | Q = Supplementary message has been provided (the message is not published due to ONS disclosure rules).                                 |
|                        | R = Recovery of price at end of sale or special offer.                                                                                  |
|                        | S = Sale price or special offer.                                                                                                        |
|                        | T = Temporarily out of stock.                                                                                                           |
|                        | W = Size change.                                                                                                                        |
| PRICE_RELATIVE_CPI     | The quote index for the individual price observation. (PRICE/BASE_PRICE_CPI).                                                           |
| PRICE_RELATIVE_RPI     | The quote index for the individual price observation. (PRICE/BASE_PRICE_RPI).                                                           |
| LOG_PRICE_RELATIVE_CPI | The natural logarithm of the individual price relative (PRICE_RELATIVE_CPI).                                                            |
| LOG_PRICE_RELATIVE_RPI | The natural logarithm of the individual price relative (PRICE_RELATIVE_RPI).                                                            |
| STRATUM_WEIGHT         | The stratum weight applicable to the item/stratum combination for the period.                                                           |
| STRATUM_TYPE           | Describes the items stratification method:                                                                                              |
|                        | 0 = Not stratified.                                                                                                                     |
|                        | 1 = Stratified by REGION.                                                                                                               |
|                        | 2 = Stratified by REGION and SHOP_TYPE.                                                                                                 |
|                        | 3 = Stratified by SHOP_TYPE.                                                                                                            |
| REGION                 | Approximation to Government Office Regions (GOR):                                                                                       |
|                        | 1 = Catalogue collections (head office).                                                                                                |
|                        | 2 = London.                                                                                                                             |
|                        | 3 = SE.                                                                                                                                 |
|                        | 4 = SW.                                                                                                                                 |
|                        | 5 = East Anglia.                                                                                                                        |
|                        | 6 = East Midlands.                                                                                                                      |
|                        | 7  = West Midlands.                                                                                                                     |
|                        | 8 = Yorks & Humber.                                                                                                                     |
|                        | 9 = NW.                                                                                                                                 |
|                        | 10 = North.                                                                                                                             |
|                        | 11 = Wales.                                                                                                                             |
|                        | 12 = Scotland.                                                                                                                          |
|                        | 13 = Northern Ireland.                                                                                                                  |
| SHOP_TYPE              | Shop type indicator:                                                                                                                    |
|                        | 1 = Multiple (10 or more outlets).                                                                                                      |
|                        | 2 = Independents (less than 10 outlets).                                                                                                |
|                        | 3 = Item not stratified by retailer.                                                                                                    |
| SHOP_WEIGHT            | The relative weight (replication factor) for the shop/item combination.                                                                 |
| BASE_PRICE_CPI         | Observed or imputed base period price (CPI).                                                                                            |
| BASE_PRICE_RPI         | Observed or imputed base period price (RPI).                                                                                            |
| STRATUM_CELL           | Stratum cell of the individual price quote. This is dependent on STRATUM_TYPE (see above):                                              |
|                        | If STRATUM_TYPE = 0, STRATUM_CELL = 0                                                                                                   |
|                        | If STRATUM_TYPE = 1, STRATUM_CELL = REGION (as derived above)                                                                           |
|                        | If STRATUM_TYPE = 3, STRATUM_CELL = SHOP_TYPE (as derived above)                                                                        |
|                        | If STRATUM_TYPE = 2 and SHOP_TYPE = 1, then STRATUM_CELL = REGION                                                                       |
|                        | If STRATUM_TYPE = 2 and SHOP_TYPE = 2, then STRATUM_CELL = REGION +13                                                                   |
| TEMPORAL               | Quotes collected on the Friday before the main index day (usually Tuesday) and excluded from the calculation of RPI.                    |
|                        | TRUE = Temporal quotes                                                                                                                  |
|                        | FALSE = Non-temporal quote                                                                                                              |
|                        | Use VALIDITY = TRUE and TEMPORAL = FALSE to filter quotes that have contributed to the construction of the RPI index.                   |
| INDEX_ALGORITHM_RPI    | The formula method used for RPI calculations (as set out in section 3.4 of the CPIH Technical Manual, 2019 edition):                    |
|                        | 1 = Dutot (RA).                                                                                                                         |
|                        | 2 = Carli (AR).                                                                                                                         |


### Raw data format (Pre Feb 2025)

| Column Heading | Description                                                                          |
|----------------|--------------------------------------------------------------------------------------|
|                |                                                                                      |
| quote_date     | year and month of data collection (yyyymm)                                           |
| item_id        | identification of the item collected                                                 |
| validity       | validity of the quote (only codes 3 and 4 used, except where indicator_box = 'N')    |
|                | 3 = Validated by system                                                              |
|                | 4 = Accepted internally by ONS staff                                                 |
| shop_code      | code of shop that the price was collected from                                       |
| price          | price of the item collected in the quote month                                       |
| indicator_box  | indicator used to highlight any item or price change                                 |
|                | S = Sale                                                                             |
|                | R = Recovery                                                                         |
|                | M = Missing                                                                          |
|                | C = Comparable                                                                       |
|                | N = Non comparable                                                                   |
|                | T = Temporarily out of stock                                                         |
|                | X = Comparable item on sale                                                          |
|                | Z = Non comparable item on sale                                                      |
| price_relative | collected price (current month) divided by base price                                |
| stratum_weight | the weight of the stratum that the price quote contributes to                        |
| stratum_type   | type of stratification used                                                          |
|                | 0 = not stratified                                                                   |
|                | 1 = region.                                                                          |
|                | 2 = region and shop                                                                  |
|                | 3 =shop                                                                              |
| end_date       | where end = 999999 price contributed to index for period otherwise price was omitted |
| region         | where the location is                                                                |
|                | 1 = Catalogue collections                                                            |
|                | 2 = London                                                                           |
|                | 3 = SE                                                                               |
|                | 4 = SW                                                                               |
|                | 5 = East Anglia                                                                      |
|                | 6 = East Midlands                                                                    |
|                | 7  = West Midlands                                                                   |
|                | 8 = Yorks & Humber                                                                   |
|                | 9 = NW                                                                               |
|                | 10 = North                                                                           |
|                | 11 = Wales                                                                           |
|                | 12 = Scotland                                                                        |
|                | 13 = NI                                                                              |
| shop_type      | either multiple or independents.                                                     |
|                | 1 = Multiple (10 or more outlets)                                                    |
|                | 2 = Independents (less than 10 outlets)                                              |
| shop_weight    | estimated number of a specific supermarket within a region                           |
| base_price     | price of an item at the base date (January of that year).                            |
| base_validity  | validity in the base month (see validity above)                                      |
| stratum_cell   | type of stratification used                                                          |
|                | There are 3 types of stratum,                                                        |
|                | by shop, by region and by region and shop.                                           |
|                | The stratum types and cells are as follows,                                          |
|                | Not stratified = 0                                                                   |
|                | By shop;                                                                             |
|                | 1 = Multiple (10 or more outlets)                                                    |
|                | 2 = Independents (less than 10 outlets)                                              |
|                | By region,                                                                           |
|                | 2 = London                                                                           |
|                | 3 = SE                                                                               |
|                | 4 = SW                                                                               |
|                | 5 = East Anglia                                                                      |
|                | 6 = East Midlands                                                                    |
|                | 7  = West Midlands                                                                   |
|                | 8 = Yorks & Humber                                                                   |
|                | 9 = NW                                                                               |
|                | 10 = North                                                                           |
|                | 11 = Wales                                                                           |
|                | 12 = Scotland                                                                        |
|                | 13 = NI                                                                              |
|                | By region and shop type,                                                             |
|                | If the shop type is a multiple, then the stratum cell is                             |
|                | equal to the regional stratum code.                                                  |
|                | If the shop type is an independent,                                                  |
|                | then the stratum cell is equal to the multiple code + 13                             |
