---
title: "ONS Shopping Price Tool data"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mm23)
```

The ONS Shopping Price tool is a new data source published in [May
2023](https://www.ons.gov.uk/economy/inflationandpriceindices/articles/shoppingpricescomparisontool/2023-05-03).
`mm23` contains some experimental functions to download and wrangle the
underlying dataset.

## Getting shopping price data

Download the latest data from ONS. `acquire_prices()` simply downloads the full
latest data download file, and returns its location. By default it puts it into a
temporary file, but you can specify where you want to put it.

As of June 2023, its not clear if the url for the data is stable.
`acquire_prices()` uses `prices_url()` for its download target. This may have to
change.

```{r}
library(mm23)

# Puts the data in a temporary file and returns its name
prices <- acquire_prices()
prices

# Or specify a location
# prices <- acquire_prices("~/data")
# data <- readxl::read_excel("~/data/datadownload.xlsx")

```

## Wrangling prices data

There are two complementary functions to access the prices data. `get_prices_metadata()` returns a dataset containing details of the data series available. `get_prices()` will return the actual data.
